```{r}
sessionInfo()
```

###-------------------------import packages-------------------------------###
```{r message=TRUE, warning=TRUE}
library("pacman")
library("BiocParallel")
library("apeglm")
library("DESeq2")
library("sva")
library("vsn")
library("dplyr")
library("stats")
library("dfidx")
library("ggplot2")
library("ggpubr")
library("ggfortify")
library("pheatmap")
library("rmarkdown")
```

###get citations for packages used
```{r}
packageList <- c("pacman", "BiocParallel", "DESeq2", "apeglm", "sva", "vsn", "dplyr", "stats",  "dfidx", "ggplot2", "ggpubr", "ggfortify", "pheatmap", "rmarkdown")
citationList <- list()

for (package in packageList){
  citationList[[package]] <- citation(package)
}

print(citationList)
```

###-----------------------set wd and import files-------------------------###
```{r}
getwd()
#setwd("D:\\Nextcloud\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis")
setwd("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis")
```

###import files
```{r}
meta_data <- read.delim("GSE121212_meta_data_collapsed_disease_type.txt", row.names = 1, header = TRUE)
#the original meta file only contained the columns "patient", "paient ID" and "condition"
#the meta data used here is a modified version of the original one
#the following columns were added "patient.id" - a continuing numbering of patients, so that each patient has a unique id
#"disease" - only the diseases -> for grouping in graphs
#"samples.type" - only the samples types, for grouping in graphs

count_matrix <- read.delim("GSE121212_readcount.txt", header = T)
#count_matrix file contains duplicates in rows -> assigning row.names = 1 here as well would result in an error
```

###------------prepare files for construction of the dds object-------------###
```{r}
#ERROR -> duplicate row names present
anyDuplicated(count_matrix$gene)       #->13535
  as.vector(count_matrix$gene[13535])   #check what the duplicate is
  count_matrix$gene[13535] = "1-Mar1"   #changed value in gene coloumn to 1-Mar1
  as.vector(count_matrix$gene[13535])   #check again if changed
anyDuplicated(count_matrix$gene)      #->13537
  as.vector(count_matrix$gene[13537])   #check what the duplicate is
  count_matrix$gene[13537] = "2-Mar1"   #changed value in gene coloumn to 2-Mar1
  as.vector(count_matrix$gene[13537])   #check again if changed
anyDuplicated(count_matrix$gene)        #->no duplicates anymore
```

###inspect data
```{r}
meta_data
nrow(meta_data)
ncol(meta_data)
dim(meta_data) #->147 x 5

head(count_matrix)
nrow(count_matrix)
ncol(count_matrix)
dim(count_matrix) #-> 31364 x 148

row.names(count_matrix) <- count_matrix$gene  #assigns "gene" column to row.names
count_matrix_new <- count_matrix[, -1]
```

#ERROR -> rownames of meta_data and colnames of count_matrix don't match 
#(some have . for _)
```{r}
head(count_matrix_new,0)
row.names(meta_data)
row.names(meta_data) <- gsub("non_lesional", "non.lesional", row.names(meta_data))
row.names(meta_data)
```

#recommended to only use _ and . as separators, changing this in the levels of factor sample.type
```{r}
meta_data$sample.type <- gsub("non-lesional", "non.lesional", meta_data$sample.type)
```
#and in the levels of factor disease
```{r}
meta_data$disease <- gsub("chronic AD", "chronic_AD", meta_data$disease)
```

#assign factors to be able to adjust settings of the analysis
```{r}
FactorList(meta_data)
meta_data$condition <- factor(meta_data$condition)
meta_data$patient.ID <- factor(meta_data$patient.ID)
meta_data$patient.id <- factor(meta_data$patient.id)
meta_data$disease <- factor(meta_data$disease)
meta_data$sample.type <- factor(meta_data$sample.type)
```

#subset meta data to remove the psorais patient, that doesnt have a paired non-lesional and lesional sample
```{r}
meta_data_subset <- meta_data
meta_data_subset <- meta_data_subset[meta_data_subset$patient.id !=  334, ]
```

#also remove the patient from the count matrix
```{r}
samples_to_exclude <- "PSO_034_lesional"
count_matrix_subset <- count_matrix_new[, !colnames(count_matrix_new) %in% samples_to_exclude]
```

#last check
```{r}
row.names(meta_data_subset)
head(count_matrix_subset, 0)
```

###--------------------constructing the DESeq2 data set (dds)---------------###
```{r}
dds <- DESeqDataSetFromMatrix(countData = count_matrix_subset,
                              colData = meta_data_subset,
                              design = ~ patient.ID + condition)
```


###----------------------------DE gene analysis-----------------------------###
```{r}
nrow(dds)
keep <- rowSums(counts(dds)) > 1 #keeps row with > 0 counts
dds <- dds[keep,]

dds$condition <- relevel(dds$condition, ref = "CTRL_healthy")

dds_res <- DESeq(dds, parallel = TRUE, BPPARAM = SnowParam(4)) #runs the DESeq2 analysis
```


###---------------removing batch effects with large sample sizes > 50-------###
```{r}
#explanation: ?svaseq()
dat <- counts(dds_res, normalized=TRUE)
idx <- rowMeans(dat) > 1
dat <- dat[idx,]
mod <- model.matrix(~ patient.ID + condition, meta_data_subset)
mod0 <- model.matrix(~ 1, colData(dds_res))
svseq <- svaseq(dat, mod, mod0, n.sv=2)

svseq$sv #shows the surrogate variables generated
```

#check how the surrogate veriable look for both the models
```{r}
par(mfrow=c(2,1),mar=c(3,5,3,1))
stripchart(svseq$sv[,1] ~ dds$patient.id,vertical=TRUE,main="SV1")
abline(h=0)
stripchart(svseq$sv[,2] ~ dds$patient.id,vertical=TRUE,main="SV2")
abline(h=0)
```

#use the sva by adding them to dds as columns
```{r}
ddssva <- dds_res
ddssva$SV1 <- svseq$sv[,1]
ddssva$SV2 <- svseq$sv[,2]

design(ddssva) <- ~ SV1 + SV2 + patient.ID + condition
```

#produce results with the new design, controlling for surrogate variables
```{r}
dds_sva_res <- DESeq(ddssva)
```

#70 rows didn't converge -> stored in mcols(dds_res)$betaconv
```{r}
mcols(dds_sva_res)$betaconv #-> says "NULL"
#these should be the rows for the empty factor levels for healthy with SV1 and SV2
#should be acceptable
```

#inspect results
```{r}
results <- results(dds_res)
results_sva <- results(dds_sva_res)

resultsNames(dds_res)
resultsNames(dds_sva_res)

summary(results)
summary(results_sva)

resOrdered <- results[order(results$padj),] #-> order results by smallest p-value increasing
head(resOrdered)

resOrdered_corrected <- results_sva[order(results_sva$padj),]
head(resOrdered_corrected)

#resOrdered_corrected["GPR183"] to check for a single entry, but this is only for one comparison
```

#less DEGs with sva,because it accounts for patient variances that and therefore identifies true DEGs
#use sva values from here!
```{r}
dds <- ddssva
dds_res <- dds_sva_res

saveRDS(dds, "dds_object.rds")
saveRDS(dds_res, "dds_res_object.rds")
```

###--------------data transformation for visualization and QC--------------###
#The point of these two transformations, the VST and the rlog, is to remove the
#dependence of the variance on the mean, particularly the high variance of the 
#logarithm of count data when the mean is low. this means, that the variance
#is higher, when there are a lot of counts (mean is high). This results in the 
#PCA being more dependent on the high count genes, which is not desired.
#Both VST and rlog use the experiment-wide trend of variance over mean, 
#in order to transform the data to remove the experiment-wide trend.
#vst = variance stabilizing transformation, recommended for n > 30.
```{r}
vsd <- vst(dds, blind=FALSE) 
vsd

meanSdPlot(assay(vsd)) #plots the SD for the data after the transformation
```

#for visualization its recommend to use shrunken values with lfcShrink()
#number 43 is from the resultsNames() -> psoriasis lesional vs healthy
```{r}
dds_shrink_PL_H <- lfcShrink(dds,
                             coef = 43,
                             type = "apeglm")
```

###---------------quality assessment and visualization---------------------###

###pheatmap
```{r}
dds <- estimateSizeFactors(dds)
#select the top 50 DEGs and store the gene names in a variable
select <- row.names(head(dds_shrink_PL_H[order(dds_shrink_PL_H$padj), ], 50))
select

#create a df used for the heatmap
df <- as.data.frame(colData(dds)[, c("disease", "sample.type")])
rownames(df) <- row.names(df)
custom_color <- list(
  sample.type = c('non.lesional' = 'lightgrey', 'lesional' = '#F44336'),
  disease = c('healthy' = 'lightgrey', 'psoriasis' = 'steelblue', 'AD' = '#FFCC80', 'chronic_AD' = '#D32F2F'))
pheatmap(assay(vsd)[select,],
         cluster_rows=TRUE, 
         cluster_cols=TRUE, 
         annotation_col=df,
         fontsize_col = 2,
         fontsize_row = 7,
         show_colnames = FALSE,
         show_rownames = TRUE,
         treeheight_row = 0,
         main = "clustering - top 50 DEGs of PSO_L_vs_H",
         annotation_colors = custom_color,
         color = colorRampPalette(c("#4A148C", "#FFF9C4", "#FF7043"))(100)
         )
#shows: roughly healthy and non-lesional samples are clustered together and 
#look rather distinct in heatmap
```
###PCA
```{r}
#collect the top 5000 gene in a variable, grouped by condition
pcaData <- plotPCA(vsd, intgroup=c("condition"), returnData = TRUE, ntop = 5000)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=condition)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  ggtitle("PCA - top 5000 genes") +
  coord_fixed() +
  #stat_ellipse(aes(group = condition), type = "norm", level = 0.95, linewidth = 1) +
  theme(
    panel.background = element_rect(fill = "transparent",
                                    colour = NA_character_), # necessary to avoid drawing panel outline
    panel.grid.major = element_blank(), # get rid of major grid
    panel.grid.minor = element_blank(), # get rid of minor grid
    plot.background = element_rect(fill = "transparent",
                                   colour = NA_character_), # necessary to avoid drawing plot outline
    legend.background = element_rect(fill = "transparent"),
    legend.key = element_rect(fill = "transparent"),
    axis.line.x.bottom=element_line(color="black"),
    axis.line.y.left=element_line(color="black"),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_color_manual(values = c("grey", "#FFB300", "#FFECB3", "#B71C1C", "#EF9A9A", "#283593", "#9FA8DA"))
 
##results quality assessment: 
#healthy and non-leasional samples are generally more similar to each other 
#than to lesional samples from the same patients
#-> this is to be expected and suggests good quality of samples
```

###------------------DEG analysis - pair-wise comparisons------------------###

#PSO
#Perform DESeq2 analysis on specified contrast
```{r}
result_PSO <- results(dds_res, contrast = c('condition', 'PSO_lesional', 'PSO_non_lesional'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
```
#filter out NAs from the padj columns as well as all padj < 0.1
```{r}
filtered_genes <- result_PSO[!is.na(result_PSO$padj) & result_PSO$padj < 0.1, ]
```
#Order and save the filtered results to a CSV file
```{r}
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
```
#write as a csv file
```{r}
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_PSO_NL_vs_PSO_L.csv"))
```
#save the object
```{r}
saveRDS(result_PSO, "DEG_Analysis_result_PSO.rds")
```

#AD
```{r}
result_AD <- results(dds_res, contrast = c('condition', 'AD_lesional', 'AD_non_lesional'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
filtered_genes <- result_AD[!is.na(result_AD$padj) & result_AD$padj < 0.1, ]
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_AD_NL_vs_AD_L.csv"))
saveRDS(result_AD, "DEG_Analysis_result_AD.rds")
```

#chronic AD
```{r}
result_CAD <- results(dds_res, contrast = c('condition', 'CAD_chronic_lesion', 'CAD_non_lesional'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
filtered_genes <- result_CAD[!is.na(result_CAD$padj) & result_CAD$padj < 0.1, ]
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_CAD_NL_vs_CAD_L.csv"))
saveRDS(result_CAD, "DEG_Analysis_result_CAD.rds")
```

#PSO L vs Healthy
```{r}
result_PSO_H <- results(dds_res, contrast = c('condition', 'PSO_lesional', 'CTRL_healthy'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
filtered_genes <- result_PSO_H[!is.na(result_PSO_H$padj) & result_PSO_H$padj < 0.1, ]
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_CTRL_H_vs_PSO_L.csv"))
saveRDS(result_PSO_H, "DEG_Analysis_result_PSO_H.rds")
```

#AD L vs Healthy
```{r}
result_AD_H <- results(dds_res, contrast = c('condition', 'AD_lesional', 'CTRL_healthy'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
filtered_genes <- result_AD_H[!is.na(result_AD_H$padj) & result_AD_H$padj < 0.1, ]
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_CTRL_H_vs_AD_L.csv"))
saveRDS(result_AD_H, "DEG_Analysis_result_AD_H.rds")
```

#CAD L vs Healthy
```{r}
result_CAD_H <- results(dds_res, contrast = c('condition', 'CAD_chronic_lesion', 'CTRL_healthy'), independentFiltering = TRUE, alpha = 0.05, pAdjustMethod = "BH", parallel = TRUE)
filtered_genes <- result_CAD_H[!is.na(result_CAD_H$padj) & result_CAD_H$padj < 0.1, ]
filtered_genes_ordered <- filtered_genes[order(filtered_genes$padj), ]
write.csv(as.data.frame(filtered_genes_ordered), row.names = TRUE, file.path("C:\\Users\\lucas\\Nextcloud\\Kurschus Lab\\Lucas\\in-silico Data\\#bulk\\GSE121212_analysis", "DEGs_CTRL_H_vs_CAD_L.csv"))
saveRDS(result_CAD_H, "DEG_Analysis_result_CAD_H.rds")
```

###-------------------------session and packacge info-----------------------###
```{r}
devtools::session_info()
```